---
title: "ESM 244 Lab 3: Spatial kriging"
author: "Casey O'Hara, Nathan Grimes, Allison Horst"
format: 
  html:
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(here)
### for working with vector spatial data:
library(sf)
### for working with rasters:
library(terra)
library(tidyterra)
### for creating cool maps:
library(tmap)
### for the geospatial statistics:
library(gstat)
library(stars)
```

# Overview

Here we will explore creating a variogram in R using groundwater depth data, and use that variogram to estimate groundwater depth in other areas. To do this, we will interpolate missing values using the kriging method.

# Read in the California depth-to-groundwater data

**Data source:** California Department of Water Resources. (2016). Groundwater Depth, California, Fall 2013. California Department of Water Resources. Available at: http://purl.stanford.edu/py212tb3061.

These data are measurements for depth to groundwater recorded in 2013. We'll read in the layer with `read_sf()`, check the CRS, then make a quick plot to start.

```{r}
### Read in the CA county data (TIGER shapefile):
ca_counties_sf <- read_sf(here("data/ca_counties"), layer = "CA_Counties_TIGER2016") %>% 
  janitor::clean_names() %>% 
  select(name)

### Read in the depth to groundwater layer: 
ca_dgw_sf <- read_sf(here('data/ca_dgw', 
                          'F2013_DBGS_Points_20150720_093252.gpkg')) %>% 
  janitor::clean_names()

```

## Explore the data

Take time to check the CRSs for each layer, transform if necessary, and then make a quick visualization.

```{r}
### Check the projection of the two layers
st_crs(ca_counties_sf) ### Pseudomercator, epsg:3857
st_crs(ca_dgw_sf)      ### WGS 84, epsg:4326

### update the groundwater data to match the counties CRS,
### and rename important variable for clarity
ca_dgw_sf <- st_transform(ca_dgw_sf, st_crs(ca_counties_sf)) %>%
  rename(depth = dgbs)

### can also tell it the EPSG code directly if you know it
# ca_dgw_sf <- st_transform(ca_dgw_sf, 3857)

### Make a quick ggplot:
ggplot() +
  geom_sf(data = ca_counties_sf) +
  geom_sf(data = ca_dgw_sf, aes(color = depth))
```

## Map these interactively to explore further

(not done in lab)

```{r}
#| eval: false ### don't include interactive map in the HTML
#| echo: true

tmap_mode("view")

tm_shape(ca_dgw_sf) +
  tm_dots("depth", palette = 'Blues')
### note - no CA map here, Leaflet does that for us

### check out all the palettes:
# tmaptools::palette_explorer() ### need package shinyjs
```

# Focus on San Joaquin County

Let's just look at depth to groundwater (ft) in and around San Joaquin County.

```{r}
### Just get sj_county
sj_county_sf <- ca_counties_sf %>% 
  filter(name == "San Joaquin")

### Now keep observations for groundwater depth within that county:
sj_depth_sf <- ca_dgw_sf %>% 
  st_intersection(sj_county_sf)

### Make a quick plot of those: 
ggplot() +
  geom_sf(data = sj_county_sf) +
  geom_sf(data = sj_depth_sf, aes(color = depth)) +
  theme_void()
```

# Spatial interpolation

## Make a variogram

Here, the "formula" specifies the dependent variable and possible covariates. Our dependent variable is depth, and we won't include any covariates, so just `depth ~ 1`.

```{r}
sj_dgw_vgm <- gstat::variogram(depth ~ 1, data = sj_depth_sf)
### what if we wanted to consider other covariates? e.g., elevation?
# sj_dgw_vgm <- gstat::variogram(depth ~ gs_elevati, data = sj_depth_sf)

### Plot it
plot(sj_dgw_vgm) 
### Looks variogramish! Increasing variance as observations get further
### apart. Let's model this...

### Now create the variogram model, & see ?vgm for other model options: 
vgm_fit <- fit.variogram(sj_dgw_vgm,
                         ### (leave with defaults first)
                         # model = vgm(model = "Gau"))
                         ### (then allow it to consider a nugget component)
                         model = vgm(model = "Gau", nugget = TRUE))

### Plot them together:
plot(sj_dgw_vgm, vgm_fit) 

### You can check the sum of squares of residuals for different models: 
attr(vgm_fit, 'SSErr') # ...and you could compare to other models...

```

## Spatial kriging

```{r}
### Make a grid over which we'll krige: 
sj_grid <- st_bbox(sj_county_sf) %>%
  stars::st_as_stars(dx = 1000, dy = 1000)

### Look at the grid we made: 
plot(sj_grid)

### Perform ordinary kriging: 
dgw_krige <- krige(depth ~ 1, sj_depth_sf, sj_grid, model = vgm_fit)

### Initial plot of kriging outcome: 
plot(dgw_krige)
```

## Making a final plot

```{r}
### Convert krige object to a raster using the terra package
krige_rast <- dgw_krige %>% 
  ### tell it the x, y, and CRS
  terra::rast() %>%
  setNames(c('pred', 'var'))

### Then we can use ggplot: 
ggplot() +
  tidyterra::geom_spatraster(data = krige_rast, aes(fill = pred)) +
  # tidyterra::geom_spatraster(data = krige_rast, aes(fill = var)) +
  geom_sf(data = sj_depth_sf, aes(fill = depth), shape = 21) +
  geom_sf(data = sj_county_sf, color = 'black', fill = NA) +
  scale_fill_gradient(low = "blue", high = "yellow")
```

We could try other models and compare how well these other models predict well depth, using cross validation - set aside a few wells out of our dataset, train the variogram on the remaining wells, run the interpolation, then see how well the kriging model predicts those set-aside wells.
